A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-03-22 CC/writeup.Rnw
small additions of examples and ggplot2 stats

\maketitle
\begin{abstract}
Abstract will be here
\end{abstract}
\hh{XXX I've set tidy to FALSE; for some reason it's totally messing up the line width.}
<<chunk_setup, echo=FALSE>>=
library(knitr)
opts_chunk$set(tidy = FALSE, echo=FALSE, eval = FALSE, message = FALSE, warning = FALSE,  fig.height = 8, fig.width = 8, fig.align = 'center')
@
<<readingdata>>=
 library(ggplot2) # needs to be the version with geom_net!
 theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
 data(blood, football, theme_elements, lesmis)
 load("data/madmen.rda") # relative path should work for both of us
names(madmen$vertices)[1] <- "label"
names(blood$vertices)[1] <- "label"

@@

\section{Introduction}
At its core, a network is simply a set of points connected in pairs by a set of lines \citep{newman}.  Here, we refer to the lines as edges and the points as vertices, although these are also called nodes.  These two seemingly simple sets of graphical objects, points and segments, are used to encode a huge variety and quantity of information across many fields of study \hh{XXX this would be a good point to cite a couple of papers that investigate networks}.  Networks vary widely in scope and complexity: the smallest network is simply an edge between two vertices, while one of the most commonly used networks, the world wide web, has billions of vertices (webpages) and billions of edges (hyperlinks) connecting them.  The edges in a network can  be directed or undirected: directed edges represent information travelling from one vertex to another, and switching the direction would change the structure of the network. The world wide web is an example of a directed network because one webpage can link to another, and not necessarily the other way around. Undirected edges, however, are simply connections between vertices.  Coauthorship networks that encode information about academic publications are examples of undirected networks because if two people author a paper together, that creates a connection between them that is bidirectional.

@@

This network is shown in figure~\ref{fig.cap:madmen_ex}. \emph{Notes: add names, color for gender}
\hh{XXX if you show the code, you should acknowledge it in the write-up. I think, for the motivating example here, it would be OK to leave out the code ... for the examples later on, the code is much more important.}

<<madmen_ex, dependson='madmen-prep', fig.cap = 'Network of Relationships in Mad Men', echo=TRUE>>=
ggplot(data = madmen$edges, aes(from_id = Name1, to_id = Name2)) +
  geom_net(vertices = madmen$vertices, vsize=3, vlabel= TRUE, ecolour="grey30",
           aes(vcolour=c( "#FF69B4", "#0099ff")[as.numeric(Gender)])) +
  expand_limits(x = c(-7.5,7.5), y = c(-7.5,7.5)) + theme_net
@

@@

 In the plot, we can see one central character who has many more relationships than any other character. This vertex represents the main character of the show, Don Draper.
\par There are many kinds of networks, and networks are extensively studied across many disciplines.  Many sociologists study social networks, and many biologists study protein networks. As different as these and the many other disciplines that study networks are, they all need the ability to quickly and effectively visualize networks.  We found the current tools to be lacking in this ability, so we chose to fill this gap by adding network plotting capabilities to the popular and widely used \texttt{R} package \texttt{ggplot2}.
\hh{XXX How widely used is ggplot2 at the moment? We could include some statistics from \url{http://cran-logs.rstudio.com/} to demonstrate this.}

@@

  \subsection{Data Structure}
  \hh{XXX we need a link between the two data frames: we have from_id and to_id in the edge data frame, but we also need an 'id' variable in the vertex data. We don't officially have that at the moment - in the code there is a soft requirement (new as of Mar 20) that one of the columns in the edge data set has to be called 'label' ... that's a bit stoopid, but I couldn't think of a better name. It would make more sense to call this id or edgeid or something along those lines - let's talk about it, and make it a formal requirement. }
  In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. \emph{The only necessary variables are the vertex names/labels and the vertex degree (if undirected) or the vertex in degree and out degree (if directed).  The names and degrees are stored in columns and the rows are the vertices.  Other values of interest, such as grouping variables, for each vertex should be stored as columns in this dataframe, with an observation for each vertex.}
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should be listed according to the names of the vertices in the vertex information data frame. Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.%\footnote{I'm not 100\% certain of all of this yet, but so far it seems to me to be the best way to think about networks in terms of data frames.}

@@

  expand_limits(x = c(-2,2), y = c(-2,2)) + theme_net #add vlabel & direction later

>>>>

\maketitle
\begin{abstract}
Abstract will be here
\end{abstract}
<<chunk_setup, echo=FALSE>>=
library(knitr)
opts_chunk$set(tidy = FALSE, echo=FALSE, eval = FALSE, message = FALSE, warning = FALSE,  fig.height = 8, fig.width = 8, fig.align = 'center')
@
<<readingdata>>=
 library(ggplot2) # needs to be the version with geom_net!
 theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
 data(blood, football, theme_elements, lesmis)
 load("data/madmen.rda") # relative path should work for both of us
names(madmen$vertices)[1] <- "label"
names(blood$vertices)[1] <- "label"

@@

\section{Introduction}
At its core, a network is simply a set of points connected in pairs by a set of lines \citep{newman}.  Here, we refer to the lines as edges and the points as vertices, although these are also called nodes.  These two seemingly simple sets of graphical objects, points and segments, are used to encode a huge variety and quantity of information across many fields of study. \sct{For instance, networks of scientific collaboration, a food web of marine animals, and American college football games are all covered in a paper on community detection in networks by \citet{football}.  Additionally, \citet{networkfailures} examine node failure in interdependent networks like power grids.  Social networks, such as links between actors found on \url{www.imdb.com}, and neural networks, like the completely mapped neural network of the \textit{C. elegans} worm are also etensively studied \citep{smallworld}.}  Networks vary widely in scope and complexity: the smallest network is simply an edge between two vertices, while one of the most commonly used networks, the world wide web, has billions of vertices (webpages) and billions of edges (hyperlinks) connecting them.  The edges in a network can  be directed or undirected: directed edges represent information travelling from one vertex to another, and switching the direction would change the structure of the network. The world wide web is an example of a directed network because one webpage can link to another, and not necessarily the other way around. Undirected edges, however, are simply connections between vertices.  Coauthorship networks that encode information about academic publications are examples of undirected networks because if two people author a paper together, that creates a connection between them that is bidirectional.

@@

This network is shown in figure~\ref{fig.cap:madmen_ex}. \emph{Notes: add names, color for gender}
<<madmen_ex, dependson='madmen-prep', fig.cap = 'Network of Relationships in Mad Men'>>=
ggplot(data = madmen$edges, aes(from_id = Name1, to_id = Name2)) +
  geom_net(vertices = madmen$vertices, vsize=3, vlabel= TRUE, ecolour="grey30",
           aes(vcolour=c( "#FF69B4", "#0099ff")[as.numeric(Gender)])) +
  expand_limits(x = c(0,1), y = c(0,1)) + theme_net
@

@@

 In the plot, we can see one central character who has many more relationships than any other character. This vertex represents the main character of the show, Don Draper.
\par There are many kinds of networks, and networks are extensively studied across many disciplines.  Many sociologists study social networks, and many biologists study protein networks. As different as these and the many other disciplines that study networks are, they all need the ability to quickly and effectively visualize networks.  We found the current tools to be lacking in this ability, so we chose to fill this gap by adding network plotting capabilities to the popular and widely used \texttt{R} package \texttt{ggplot2}. \sct{From January 1, 2015 to March 21, 2015, \texttt{ggplot2} was downloaded over 270,000 times: approximately 3454 downloads per day. It has also been downloaded in 189 countries at least once, and in 31 of those countries, including China, Israel, and Colombia, it has been downloaded over 1,000 times.}

@@

  \subsection{Data Structure}
  \hh{XXX we need a link between the two data frames: we have from\_id and to\_id in the edge data frame, but we also need an 'id' variable in the vertex data. We don't officially have that at the moment - in the code there is a soft requirement (new as of Mar 20) that one of the columns in the edge data set has to be called 'label' ... that's a bit stoopid, but I couldn't think of a better name. It would make more sense to call this id or edgeid or something along those lines - let's talk about it, and make it a formal requirement.}
  In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. \emph{The only necessary variables are the vertex names/labels and the vertex degree (if undirected) or the vertex in degree and out degree (if directed).  The names and degrees are stored in columns and the rows are the vertices.  Other values of interest, such as grouping variables, for each vertex should be stored as columns in this dataframe, with an observation for each vertex.}
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should be listed according to the names of the vertices in the vertex information data frame. Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.%\footnote{I'm not 100\% certain of all of this yet, but so far it seems to me to be the best way to think about networks in terms of data frames.}

@@

  expand_limits(x = c(0,1), y = c(0,1)) + theme_net #add vlabel & direction later
